Learning complex models with invertible neural networks: a likelihood-free Bayesian approach

### 2019-06-24 latex/paper_main.tex
new latex files

\dropcap{M}athematical models are formal descriptions of scientific theories allowing a clear and unambiguous way to formulate and test scientific hypotheses about probabilistic phenomena in a probabilistic world. In its most abstract form, a mathematical model is specified by a set of parameters $\boldsymbol{\theta}$ and a forward model $q$ mimicking the process by which manifest data $\boldsymbol{x}$ arise from latent parameters: 
\begin{align*}
\boldsymbol{x}= q(\boldsymbol{\theta}) \numberthis \label{eqn:1} 
\end{align*}
While $q$ can represent an arbitrarily complex process by an arbitrarily complicated expression, its functional form is usually guided by a well-founded theoretical framework. For instance, $q$ can be a stochastic differential equation describing the dynamics of single neurons in the brain, or a step-by-step biological algorithm dictating the rate of gene expression in certain cells. Thus, it is only through theoretical embedding that a meaningful interpretation in terms of some mechanism can be attached to the parameters of a mathematical model. Examples of mathematical models can be found in various scientific domains, e.g., genetics \cite{zappia2017splatter, beaumont2002approximate}, cognitive science \cite{palestro2018likelihood, usher2001time}, neuroscience \cite{hwang2018conditional, lueckmann2017flexible}, population dynamics \cite{wood2010statistical}, epidemiology \cite{hethcote2000mathematics}, just to name a few.

@@

The \textit{invertible network} is responsible for learning the posterior of the model parameters given the observed and summarized data. It is based on the recently developed flow-based architecture \cite{kingma2018glow, grover2018flow, dinh2016density}. Flow-based methods provide exact latent-variable inference and log-likelihood evaluation when operating at optimum. In the \textbf{Methods} section, we show that our method maximizes the posterior over model parameters directly when cast in the context of likelihood-free inference. Furthermore, flow-based methods are capable of approximating very high-dimensional distributions (e.g., images). Once trained with a sufficient amount of simulated data, our invertible network can perform full Bayesian inference on large amounts of real data from a given domain in a single pass.

@@

\subsection*{Example 2 - The Lévy-Flight Model}
Evidence accumulator models (EAMs) describe (perceptual) decision making by a set of neurocognitively motivated parameters CITE. EAMs are most often applied to choice reaction times (RT) data to obtain an estimate of the neurocognitive processes governing observed RT distributions in human and animal participants. Most EAM flavors share four underlying assumptions: \textit{i)} information about a stimulus (response option) is accumulated continuously through time; \textit{ii)} stochasticity in the form of noisy accumulation ensures variability; \textit{iii)} empirical response times can be decomposed into a decision time component and a non-decision time component accounting for pre-decisional perceptual (encoding time) and post-decisional motor processes (response execution); and \textit{iv)} a decision is met when the activation of an accumulator exceeds a threshold. In its most general formulation, the forward model of EAMs takes the form of a stochastic differential equation given by:
\begin{align*} 
dx = vdt + cd\xi  \numberthis \label{eqn:stillno}
\end{align*}
where $dx$ denotes a change in activation of an accumulator, $v$ denotes the average speed of information processing (often termed the drift rate), and $d\xi$ represents a stochastic additive component, usually modeled as following a Gaussian distribution centered around $0$: $d\xi \sim \mathcal{N}(0, c^{2})$.

EAMs are particularly amenable for likelihood-free inference, since most members of the family turn out to be intractable. This intractability has precluded many interesting applications and empirically driven model refinements. Here, we apply our method to estimate the parameters of the recently proposed Lévy-Flight Model (LFM, cite). The LFM assumes an \textit{alpha-stable} noise distribution of the evidence accumulation process in order to model "jumps" in the decision process. The LFM does not have a known likelihood and a fully Bayesian treatment of the model is still missing from the literature. The forward equation of the LFM is given by:
\begin{align*}
dx &= vdt + \xi dt^{1/\alpha} \numberthis \label{eqn:stillno2} \\
\xi &\sim AlphaStable(\alpha,0,1,0) \numberthis \label{eqn:stillno3}
\end{align*}
The LFM has three additional parameters: a threshold $a$ determining the amount of evidence needed for the termination of a decision process; a relative starting point, $zr$, determining the amount of starting evidence available to the accumulator before the actual decision alternatives are presented; and an additive non-decision time $t_{0}$.

@@

\subsection*{Example 3 – The Stochastic SIR Model}
Compartmental models in epidemiology are used to describe the stochastic dynamics of infectious diseases as they spread over a population of individuals \cite{hethcote2000mathematics}. The parameters of compartmental models encode important characteristics of diseases, such as the rates of infection or recovery from the disease. The stochastic SIR model describes the transition dynamics of N individuals between three discrete states: susceptible (S), infected (I), and recovered (R). The transition dynamics are given by the following equations:
\begin{align*}
\triangle S &= -\triangle N_{SI} \numberthis \\
\triangle I &= \triangle N_{SI} - \triangle N_{IR} \numberthis \\
\triangle R &= \triangle N_{IR} \numberthis \\
\triangle N_{SI} &\sim Binomial(S, 1 - \exp\left(-\beta\frac{I}{N}\triangle t \right)) \numberthis \\
\triangle N_{IR} &\sim Binomial(I, 1 - \exp\left(-\gamma\triangle t \right)) \numberthis 
\end{align*}
where $S + I + R = N$ give the number of susceptible, infected, and recovered individuals, respectively. The parameter $\beta$ controls the transition rate from being susceptible to infected, and $\gamma$ controls the transition rate from being infected to recovered. The number of individuals moving from $S$ to $I$, given by $\triangle N_{SI}$, and the number of people moving from $I$ to $R$, given by $\triangle N_{IR}$, over a time interval $\triangle t$ are modeled as binomial random variables. The above listed stochastic system has no known analytic solution and thus requires numerical simulation methods for finding optimal parameter values. Cast as a parameter estimation task, the challenge is to recover $\boldsymbol{\theta} = \{\beta,\gamma\}$ from three dimensional time-series data $\boldsymbol{X} = (\boldsymbol{x}_{1},\boldsymbol{x}_{2},...,\boldsymbol{x}_{T})$ where each $\boldsymbol{x}_{t} \in \mathbb{N}^{3}$ is a triple containing the number of susceptible ($S$), number of infected ($I$), and recovered ($R$) individuals at time $t$.

>>>>

\dropcap{M}athematical models are formal descriptions of scientific theories allowing a clear and unambiguous way to formulate and test scientific hypotheses about probabilistic phenomena in a probabilistic world. In its most abstract form, a mathematical model is specified by a set of parameters $\boldsymbol{\theta}$ and a forward model $q$ mimicking the process by which manifest data $\boldsymbol{x}$ arise from latent parameters: 
\begin{align*}
\boldsymbol{x}= q(\boldsymbol{\theta}) \numberthis \label{eqn:1} 
\end{align*}
While $q$ can represent an arbitrarily complex process by an arbitrarily complicated expression, its functional form is usually guided by a well-founded theoretical framework. For instance, $q$ can be a stochastic differential equation describing the dynamics of single neurons in the brain, or a step-by-step biological algorithm dictating the rate of gene expression in certain cells. Thus, it is only through theoretical embedding that a meaningful interpretation in terms of some mechanism can be attached to the parameters of a mathematical model. Examples of mathematical models can be found in various scientific domains, e.g., genetics \cite{zappia2017splatter, beaumont2002approximate}, cognitive science \cite{palestro2018likelihood, usher2001time}, neuroscience \cite{hwang2018conditional, lueckmann2017flexible}, population dynamics \cite{wood2010statistical}, epidemiology \cite{keeling2011modeling,hethcote2000mathematics}, just to name a few.

@@

The \textit{invertible network} is responsible for learning the posterior of the model parameters given the observed and summarized data. It is based on the recently developed flow-based architecture \cite{kingma2018glow, grover2018flow, dinh2016density}. Flow-based methods provide exact latent-variable inference and log-likelihood evaluation when operating at optimum. In the \textbf{Methods} section, we show that our method maximizes the posterior over model parameters directly when cast in the context of likelihood-free inference. Furthermore, flow-based methods are capable of approximating very high-dimensional distributions (e.g., the pixels of an image). Once trained with a sufficient amount of simulated data, our invertible network can perform full Bayesian inference on large amounts of real data from a given domain in a single pass.

@@

\subsection*{Example 2 - The Lévy-Flight Model}
Evidence accumulator models (EAMs) describe (perceptual) decision making by a set of neurocognitively motivated parameters \cite{ratcliff2008diffusion}. EAMs are most often applied to choice reaction times (RT) data to obtain an estimate of the neurocognitive processes governing observed RT distributions in human and animal participants. Most EAM variants share four underlying assumptions: \textit{i)} information about a stimulus (response option) is accumulated continuously through time; \textit{ii)} stochasticity in the form of noisy accumulation ensures variability; \textit{iii)} empirical response times can be decomposed into a decision time component and a non-decision time component accounting for pre-decisional perceptual (encoding time) and post-decisional motor processes (response execution); and \textit{iv)} a decision is met when the activation of an accumulator exceeds a threshold. In its most general formulation, the forward model of EAMs takes the form of a stochastic differential equation given by \cite{usher2001time}:
\begin{align*} 
dx = vdt + cd\xi  \numberthis \label{eqn:stillno}
\end{align*}
where $dx$ denotes a change in activation of an accumulator, $v$ denotes the average speed of information processing (often termed the drift rate), and $d\xi$ represents a stochastic additive component, usually modeled as following a Gaussian distribution centered around $0$: $d\xi \sim \mathcal{N}(0, c^{2})$.

EAMs are particularly amenable for likelihood-free inference, since most members of the family turn out to be intractable \cite{miletic2017parameter}. This intractability has precluded many interesting applications and empirically driven model refinements. Here, we apply our method to estimate the parameters of the recently proposed Lévy-Flight Model (LFM, \cite{voss2019sequential}). The LFM assumes an \textit{alpha-stable} noise distribution of the evidence accumulation process in order to model "jumps" in the decision process. However, the inclusion of \textit{alpha-stable} noise leads to a model with intractable likelihood; further, to our knowledge, a fully Bayesian treatment of the model is still missing from the literature. The forward equation of the LFM is given by:
\begin{align*}
dx &= vdt + \xi dt^{1/\alpha} \numberthis \label{eqn:stillno2} \\
\xi &\sim AlphaStable(\alpha,0,1,0) \numberthis \label{eqn:stillno3}
\end{align*}
The LFM has three additional parameters: a threshold $a$ determining the amount of evidence needed for the termination of a decision process; a relative starting point, $zr$, determining the amount of starting evidence available to the accumulator before the actual decision alternatives are presented; and an additive non-decision time $t_{0}$.

@@

\subsection*{Example 3 – The Stochastic SIR Model}
Compartmental models in epidemiology are used to describe the stochastic dynamics of infectious diseases as they spread over a population of individuals \cite{sahneh2017gemfsim,keeling2011modeling, hethcote2000mathematics}. The parameters of compartmental models encode important characteristics of diseases, such as the rates of infection or recovery from the disease. The stochastic SIR model describes the transition dynamics of $N$ individuals between three discrete states: susceptible ($S$), infected ($I$), and recovered ($R$). The transition dynamics are given by the following equations:
\begin{align*}
\triangle S &= -\triangle N_{SI} \numberthis \\
\triangle I &= \triangle N_{SI} - \triangle N_{IR} \numberthis \\
\triangle R &= \triangle N_{IR} \numberthis \\
\triangle N_{SI} &\sim Binomial(S, 1 - \exp\left(-\beta\frac{I}{N}\triangle t \right)) \numberthis \\
\triangle N_{IR} &\sim Binomial(I, 1 - \exp\left(-\gamma\triangle t \right)) \numberthis 
\end{align*}
where $S + I + R = N$ give the number of susceptible, infected, and recovered individuals, respectively. The parameter $\beta$ controls the transition rate from being susceptible to infected, and $\gamma$ controls the transition rate from being infected to recovered. The number of individuals moving from $S$ to $I$, given by $\triangle N_{SI}$, and the number of people moving from $I$ to $R$, given by $\triangle N_{IR}$, over a time interval $\triangle t$ are modeled as binomial random variables. The above listed stochastic system has no known analytic solution and thus requires numerical simulation methods for finding optimal parameter values. Cast as a parameter estimation task, the challenge is to recover $\boldsymbol{\theta} = \{\beta,\gamma\}$ from three dimensional time-series data $\boldsymbol{X} = (\boldsymbol{x}_{1},\boldsymbol{x}_{2},...,\boldsymbol{x}_{T})$ where each $\boldsymbol{x}_{t} \in \mathbb{N}^{3}$ is a triple containing the number of susceptible ($S$), number of infected ($I$), and recovered ($R$) individuals at time $t$.
